---
title: "R Notebook"
output: html_notebook
---

# Planning my attack
When I plan my analyses, I usually do it in in this order: First, I think about what my *research question* is. Second, I think about what I need to *meausure*, third, I think about how I will *communicate* that to my audience. I can do all of this before I have data, and doing this helps me to stay on track. Everything I do must help me to reach this goal.

For this tutorial: First, I want to know what the relationship is between race and the incidence of COVID-19. Second, I think that I will measure this by collecting county-level data for the percent black, and fraction of people with COVID-19. Third, I think that I want to visualize this with maps of percent black and percent with COVID, and perhaps with a correlation coefficient between the two. In this tutorial, though, I'll just do the mapping.

How will I do this? I'll need the following data:

Percent Black or African-American for each county.  Where will I get this? The Census Bureau. I might find Percent Black, but I might also only find the Population that is Black or African American, in which case I'll also need the Total Population so that I calculate the Percent.

I'll also need the Percent of people in each county that have had COVID-19. Where will I get this? If I am working with a single state, I might go to the State Department of Health, and see if they have data to download. If I am working with the entire nation, I might go to someplase like the Johns Hopkins University dataset that has done a lot of work to collect data from all 50 states and to put them in one file. Again, I might not find the percent, so I might need to get the number of total cases and the Total Population.

I'll also need some way to map the data.  I need data with the name and boundary of each county.

Also, all of these data are going to have be joined together. So that's something to keep an eye out for.

# Getting the data
## Getting the race data
The Census Bureau is the ultimate source of all official data in the United States about demographic characteristics like Race.  But there is a helpful R package to get this, called `tidycensus`. So make sure that you have the package first by running the command `install.packages('tidycensus')`.

The tidycensus package requires you to have a Census API key.  Anybody can get one, but it will take you a few minutes. https://api.census.gov/data/key_signup.html

Load the library and enter your api key
```{r}
library(tidycensus)
# Uncomment the next line and paste in an api key
# census_api_key('KEY GOES HERE')
```

There is a ton of census data, and you need to look through a catalog in order to find what you want. 
Load the catalog for the 2018 American Community Survey Demographic Profile
```{r}
acs_catalog <- load_variables(year = 2018, dataset='acs5/profile')
```

There's only 1,346 variables to look through. In the command console, type `View(acs_variables)` and then search for the word "Black". Fortunately, there are only a few. There's even one for percent! I recommend using called "DP05_0065P".

Looking ahead to other analyses we might do, I'm also going to get Percent Hispanic ("DP05_0070P"), Percent in Poverty ("DP03_0119P"), Median Age ("DP05__0018") and Total Population ("DP05_001")

```{r}
county_census <- get_acs(geography = 'county',
                         variables = c('DP05_0001','DP05_0018','DP05_0065P','DP05_0071P','DP05_0018','DP03_0119P'),
                         state = 'TN',
                         output = 'wide')
county_census
```

I'm going to rename things so it's not as gross
```{r}
county_census <- select(county_census, GEOID, NAME,
                        population = DP05_0001E,
                        age = DP05_0018E,
                        pct_black = DP05_0065PE,
                        pct_hisp = DP05_0071PE,
                        pct_poverty = DP03_0119PE
                        )
county_census
```

That's much better. We're done with getting the census data.


## Get the COVID data.


