---
title: "COVID 19 package"
output: html_document
---

A group of scholars in Italy has created a really neat package for downloading a bunch of COVID data all at once, for the entire world.

```{r}
library(tidyverse)
library(COVID19)
```


The following code will download county level data for the US.
```{r}
covid_county <- covid19(c('US'), level=3)
```
To getstate-level data, change level to 2.

There's a ton of stuff here:

```{r}
glimpse(covid_county)
```

Note the `key_numeric` column: that is a FIPS code that you can match to the GEOID in the Census shapefiles. Unfortunately, not every state is present.

Here's a quick map of deaths on April 15th. We'll use hollow bubbles at each lat/lon. Note, this is not a shapefile, so we can't use geom_sf. This is a geom_point(): I've used shape=1 to tell it to use bubbles (I had to google this) and alpha=.5 to allow the circles to be slightly transparent, and I'll make all the circles blue because who doesn't love blue?

```{r}
data <- covid_county %>%
  filter(latitude > 26) %>% # Sorry Hawaii
  filter(latitude < 50) # Sorry Alaska
ggplot(data = data %>% filter(date == lubridate::as_date('2020-04-15')),
       mapping = aes(x=longitude, y=latitude)) + 
  geom_point(aes(size = deaths), alpha=.5, shape=1, color='blue') +
  coord_map() # Tells ggplot to plot with the mercator projection
```

I want to get rid of the counties with zero counts, because I think that they are misleading. Also, mercator maps are ugly:

```{r}
ggplot(data = data %>% 
         filter(date == lubridate::as_date('2020-04-15')) %>%
         filter(deaths > 0),
       mapping = aes(x=longitude, y=latitude)) + 
  geom_point(aes(size = deaths), alpha=.5, shape=1, color='blue') +
  scale_size_area(max_size=8) + # Adjust the maximum size if necessary.
  coord_map("albers",lat0=39, lat1=45) # Tells ggplot to plot with the Albers projection, centered over the United States.
```

It would be great to have a boundary map of states.  I had to google this, but his page helped: https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html

```{r}
library(ggmap)
state_boundaries <- map_data('state')

ggplot(data = data %>% 
         filter(date == lubridate::as_date('2020-05-15')) %>%
         filter(deaths>0),
       mapping = aes(x=longitude, y=latitude)) + 
  geom_path(data = state_boundaries,
               mapping = aes(x=long, y=lat, group=group)) +
  geom_point(aes(size = deaths), alpha=.5, shape=1, color='blue') +
  scale_size_area(max_size=8) +
  coord_map("albers",lat0=39, lat1=45)
```
Now I want to lighten up the state boundaries, and remove the grid:

```{r}
ggplot(data = data %>% 
         filter(date == lubridate::as_date('2020-05-15')) %>%
         filter(deaths>0),
       mapping = aes(x=longitude, y=latitude)) + 
  geom_path(data = state_boundaries,
               mapping = aes(x=long, y=lat, group=group), color='grey60') +
  geom_point(aes(size = deaths), alpha=.5, shape=1, color='blue') +
  scale_size_area(max_size=8) + 
  coord_map("albers",lat0=39, lat1=45) + 
theme_void()
```





Finally, Let's create a time series of plots. I think this is pretty cool. It requires a different filter of the data, but the map time series is just one extral line of code:  adding a `facet_wrap()` like so:


```{r}
dates <- c('2020-03-01','2020-03-15','2020-04-01','2020-04-15','2020-05-01','2020-05-15','2020-06-01')
ggplot(data = data %>% 
         filter(date %in% lubridate::as_date(dates)) %>%
         filter(deaths>0),
       mapping = aes(x=longitude, y=latitude)) + 
  geom_path(data = state_boundaries,
               mapping = aes(x=long, y=lat, group=group), color='grey60') +
  geom_point(aes(size = deaths), alpha=.5, shape=1, color='blue') +
  scale_size_area(max_size=8) +
  facet_wrap(~date) +
  coord_map("albers",lat0=39, lat1=45) +
  theme_void()
```

This will be best saved out as a full size image: width = 11 and height = 8.5.


Finally, here's some code that I used to experiment with merging it to the county shapefile. Not every county could be merged, and it wasn't very attractive. But I'm putting it here for study.

```{r}
library(sf)
county_map <- st_read('data/cb_2018_us_county_20m.shp')
county_map <- left_join(
  x = county_map %>% mutate(GEOID = as.character(GEOID)),
  y = covid_county %>% mutate(GEOID = as.character(key_numeric)),
  by=c('GEOID'))

plt <- ggplot(data = county_map %>% 
         filter(date == lubridate::as_date('2020-04-15'))) + 
  geom_sf(mapping = aes(fill = deaths))
```



