---
title: "lab02_key"
author: "Nicholas Nagle"
date: "9/5/2019"
output:
  word_document: default
  html_document: default
---

```{r}
library(tidyverse)
library(sf)
```

Load the shapefile 'knox_acs.shp' into R.

```{r}
knox <- st_read('knox_acs.shp')
```

Create a new column called 'college_frac' that has the percentage of population between ages 18 and 21.

```{r}
knox <- knox  %>%
  mutate(college_frac = AGE18_21 / TOTAL_POP)
```

Create a choropleth map of the proportion of the population age 18-21.
```{r}
ggplot(knox) + ggspatial::annotation_map_tile() +
  geom_sf(aes(fill=college_frac), alpha=.7) + 
  labs(fill = "Proportion age 18-21")
```

Plot a histogram of Median HH Income and label the axes appropriately. Describe the shape of this distribution.

```{r}
ggplot(knox) + geom_histogram(aes(x=MEDHHINC/1000), binwidth=10) + labs(x='Median HH Income ($1000s)')

```


Create a scatterplot with HHincome (in 1000s) on the x axis and House Value (in $1000s) on the y-axis. Set the alpha level to the population of each census tract.

```{r, fig.width=5, fig.height=5, dev='pdf'}
ggplot(knox, aes(x=MEDHHINC/1000,y=MEDHVALUE/1000, alpha=TOTAL_POP)) + geom_point() +
  labs(x='Median HH Income ($1000s)', y='Median House Value ($1000s)', alpha='Population')
```

Calculate the three correlation coefficients between HH Income, House Value, and percent 18-21.
Display those correlation coefficients as a table AND use them in markdown in a sentence.

Briefly explain why you think each correlation is strong or weak, positive or negative.


```{r}
summary_stat <- knox %>% st_set_geometry(NULL) %>% 
  summarize(mn_inc = mean(MEDHHINC, na.rm=TRUE),
            cor1 = cor(MEDHHINC, MEDHVALUE,use='complete.obs'),
            cor2 = cor(MEDHHINC, college_frac, use='complete.obs'),
            cor3 = cor(MEDHVALUE, college_frac, use='complete.obs')) 

summary_stat$cor1
```

The correlation between HH income and Home value is `r summary_stat$cor1`.
The correlation between HH income and % ages 18-21 is `r summary_stat$cor2`.
The correlation between home value and % ages 18-21 is `r summary_stat$cor3`.

```{r}
summary_stat <- knox %>% st_set_geometry(NULL) %>% 
  summarize(mn_inc = mean(MEDHHINC, na.rm=TRUE),
            cor1 = cor(MEDHHINC, MEDHVALUE,use='complete.obs'),
            cor2 = cor(MEDHHINC, college_frac, use='complete.obs'),
            cor3 = cor(MEDHVALUE, college_frac, use='complete.obs')) 


```



The correlationis `r summary_stat$cor1`







